---
title: "Noise_data_preparation"
author: "Tino Schneidewind"
date: "2025-05-09"
output: html_document
---

Amended in July 2026 by Apolline Saucy

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Aim of the document

In this RMarkdown I will implement the data preparation of the noise sentry data for the analysis following the steps provided by Apolline Saucy. For this trial, I will utilise only noise data from one participant of the first week (001).

```{r load libs, echo=FALSE, warning=FALSE, message=FALSE, fig.height=4, fig.width=8, fig.align='center'}
rm(list=ls())


# for handling file paths and different operating systems
source("../functions.R")
# libraries
library(readr);library(tidyr);library(dplyr);library(readxl);library(zoo)
library(lubridate);library(stringr);library(ggplot2);library(gridExtra); library(grid)

# specify the week to compile (needs to match naming convention on synology)
week_indicator = "week_1"



# LOAD DATA
#---- 

# load redcap from CCH
  # iButton and Noise data
  data <- read_csv("/Volumes/FS/_ISPM/CCH/Actual_Project/data-raw/Participants/week_1_IB_RAW_data_unclean.csv")
  # REDCap for uids and start and end times
  redcap = read_csv("/Volumes/FS/_ISPM/CCH/Actual_Project/data/App_Personal_Data_Screening/redcap_all.csv")
  

# select redcap data 
redcap <- redcap |> 
  select(uid, redcap_event_name, pvl_start, pvl_end, starts_with("pvl_ib")) |>
  drop_na(pvl_start) |>
  filter(redcap_event_name == "study_visit_week_1_arm_1") |>
  filter(!(uid %in% c("ACT029U", "ACT034X", "ACT045O")))

redcap_001 <- redcap |>
  filter(uid == "ACT001D") |>
  mutate(dow_start = weekdays(pvl_start),
         dow_end = weekdays(pvl_end))


# noise data
data_N <- data |>
  mutate(Variable = if_else(Variable == "_NS", "NS", Variable)) |>
  filter(Variable == "NS")|>
  pivot_wider(names_from = Variable, values_from = Value)|>
  mutate(
    NS_MA = rollmean(NS, k = 8, fill = NA, align = "left"),
    hour = hour(datetime),
    dow = wday(datetime, week_start = 1))


# noise data for a specific participant
data_N_001 <- data_N |>
  filter(uid == "ACT001D") 

head(data_N_001)

ggplot(data_N_001, aes(x = datetime)) +
  geom_line(aes(y = NS)) +
  geom_point(aes(y = NS)) +
  labs(x = "time", y = "NS [dB]", title = "Uncleaned noise data"
  ) +
  theme_minimal()

```

<br>

##### 1) Clean the raw data 

as described in [Land Use Regression Modeling of Outdoor Noise Exposure in Informal Settlements in Western Cape, South Africa](https://www.mdpi.com/1660-4601/14/10/1262)

a) restrict to the same weekly period for everyone. Identify start and end times for which everyone is being monitored. Time windows can be discussed based on data availability.

Tuesday 6 am to saturday 6 am

```{r 1a}
data_N_001 <- data_N_001 |>
  filter(
    (dow == 2 & hour >= 6) |  # Tuesday from 6am
      (dow %in% 3:5) |        # Tuesday to Thursday
      (dow == 6 & hour < 6) # Saturday up to 6am
  )
```

b)
Exclude observations with >10% missing data (threshold to discuss depending on data availability)

```{r 1b}
NA_perc = mean(is.na(data_N_001$NS))

if(NA_perc > .1){
  data_N_001$NS = NA
}
```


c)
Remove outliers, defined as one-minute noise measurements (=raw data) exceeding the five-day mean by plus or minus three standard deviations.

```{r 1c}
mean_noise = mean(data_N_001$NS, na.rm = TRUE)
sd_noise = sd(data_N_001$NS, na.rm = TRUE) * 3

data_N_001 <- data_N_001 |>
  mutate(
    NS_outliers = ifelse(NS > mean_noise + sd_noise | NS < mean_noise - sd_noise, TRUE, FALSE)
  )
```

```{r 1c plot, echo=FALSE, fig.height=4, fig.width=8, fig.align='center'}
ggplot(data_N_001, aes(x = datetime)) +
  geom_line(aes(y = NS)) +
  geom_point(aes(y = NS, color = as.factor(NS_outliers))) +
  scale_color_manual(values = c("FALSE" = "black", "TRUE" = "red")) +
  labs(x = "time", y = "NS [dB]", title = "Uncleaned Noise Data", subtitle = "with outliers", color = "Outlier") +
  theme_minimal()
```


<br>

##### 2) Calculate noise indicators (long-term) 

A-weighted equivalent sound level variables as described here: [Space-time characterization of community noise and sound sources in Accra, Ghana | Scientific Reports](https://www.nature.com/articles/s41598-021-90454-6#Fig1)

```{r 2}
data_N_001 <- data_N_001 |>
  mutate(
    NS_clean = ifelse(NS_outliers == TRUE, NA, NS)
  )
```

a) calculate L_Aeq and intermittency ratios for each site and date of measurement

```{r 2a}
# total
L_Aeq_tot = 10 * log10( mean(10^(0.1 * data_N_001$NS_clean), na.rm = TRUE ) ) 

# hourly
noise_ind_hourly <- data_N_001 |>
  group_by(dow, hour) |>
  
  # A-weighted equivalent continuous hourly sound levels L_Aeq_1h
  summarize(L_Aeq_1h = 10 * log10( mean(10^(0.1 * NS_clean), na.rm = TRUE ) ),
            .groups = "drop")
```


```{r 2aplot, echo=FALSE, fig.height=4, fig.width=8, fig.align='center'}
ggplot(noise_ind_hourly, aes(y = L_Aeq_1h, x = 1:length(L_Aeq_1h))) +
  geom_line() + 
  labs(x = "hours since Monday 8 am") +
  theme_minimal() +
  lims(y = c(35,75))
```


```{r 2a daily}
# daily data frame
noise_ind_daily <- data_N_001 |>
  group_by(dow) |>
  
      mutate(
      energy = 10^(0.1 * NS) # work on energy level (linear) before going back to decibel scale
    )|>
    summarize( 
      
      # LAeq --------------------
      L_Aeq_tot = 10 * log10(mean(energy, na.rm = TRUE)),
      L_Aeq_day = 10 * log10(mean(energy[hour >= 6 & hour < 22], na.rm = TRUE)),
      L_Aeq_night = 10 * log10(mean(energy[hour < 6 | hour >= 22], na.rm = TRUE)),
      
      # IR 24h --------------------
      IR_24h = {
        Ltot <- 10 * log10(mean(energy, na.rm = TRUE))
        K <- Ltot + 3 # threshold is average sound level + 3
        idx <- NS >= K # identify observations which overpass threshold K
        sum(energy[idx], na.rm = TRUE) / sum(energy, na.rm = TRUE) # IR is the proportion of overall energy of the events to the overall weekly energy
      },
    
      # IR day ------------------
      IR_day = {
        sel <- hour >= 6 & hour < 22
        x_energy <- energy[sel]
        x_NS <- NS[sel]
        Ltot <- 10 * log10(mean(x_energy, na.rm = TRUE))
        K <- Ltot + 3
        idx <- x_NS >= K
        sum(x_energy[idx], na.rm = TRUE) / sum(x_energy, na.rm = TRUE)
      },
      
      # IR night ------------------
      IR_night = {
        sel <- hour < 6 | hour >= 22
        x_energy <- energy[sel]
        x_NS <- NS[sel]
        Ltot <- 10 * log10(mean(x_energy, na.rm = TRUE))
        K <- Ltot + 3
        idx <- x_NS >= K
        sum(x_energy[idx], na.rm = TRUE) / sum(x_energy, na.rm = TRUE)
      },
      
      # L_den ------------------
      L_den_day = 10 * log10(mean(energy[hour >= 6 & hour < 18], na.rm = TRUE)),
      L_den_eve = 10 * log10(mean(energy[hour >= 18 & hour < 22], na.rm = TRUE)),
      L_den_nig = 10 * log10(mean(energy[hour < 6 | hour >= 22], na.rm = TRUE)),
      
      L_den = 10 * log10(
        (1/24) *
          (12 * 10^(L_den_day / 10) +
             4  * 10^((L_den_eve + 5) / 10) +
             8  * 10^((L_den_nig + 10) / 10))
      ),
      
      .groups = "drop"
)


# weekly data frame
noise_ind_weekly <- data_N_001 |>
      mutate(
      energy = 10^(0.1 * NS) # work on energy level (linear) before going back to decibel scale
    )|>
    summarize( 
      
      # LAeq --------------------
      L_Aeq_tot = 10 * log10(mean(energy, na.rm = TRUE)),
      L_Aeq_day = 10 * log10(mean(energy[hour >= 6 & hour < 22], na.rm = TRUE)),
      L_Aeq_night = 10 * log10(mean(energy[hour < 6 | hour >= 22], na.rm = TRUE)),
      
      # IR 24h --------------------
      IR_24h = {
        Ltot <- 10 * log10(mean(energy, na.rm = TRUE))
        K <- Ltot + 3 # threshold is average sound level + 3
        idx <- NS >= K # identify observations which overpass threshold K
        sum(energy[idx], na.rm = TRUE) / sum(energy, na.rm = TRUE) # IR is the proportion of overall energy of the events to the overall weekly energy
      },
    
      # IR day ------------------
      IR_day = {
        sel <- hour >= 6 & hour < 22
        x_energy <- energy[sel]
        x_NS <- NS[sel]
        Ltot <- 10 * log10(mean(x_energy, na.rm = TRUE))
        K <- Ltot + 3
        idx <- x_NS >= K
        sum(x_energy[idx], na.rm = TRUE) / sum(x_energy, na.rm = TRUE)
      },
      
      # IR night ------------------
      IR_night = {
        sel <- hour < 6 | hour >= 22
        x_energy <- energy[sel]
        x_NS <- NS[sel]
        Ltot <- 10 * log10(mean(x_energy, na.rm = TRUE))
        K <- Ltot + 3
        idx <- x_NS >= K
        sum(x_energy[idx], na.rm = TRUE) / sum(x_energy, na.rm = TRUE)
      },
      
      # L_den ------------------
      L_den_day = 10 * log10(mean(energy[hour >= 6 & hour < 18], na.rm = TRUE)),
      L_den_eve = 10 * log10(mean(energy[hour >= 18 & hour < 22], na.rm = TRUE)),
      L_den_nig = 10 * log10(mean(energy[hour < 6 | hour >= 22], na.rm = TRUE)),
      
      L_den = 10 * log10(
        (1/24) *
          (12 * 10^(L_den_day / 10) +
             4  * 10^((L_den_eve + 5) / 10) +
             8  * 10^((L_den_nig + 10) / 10))
      ),
      
      .groups = "drop"
    )
```


##### 3) Additional cleaning step: clean hourly averages {.tabset}
Hourly averages (LAeq_hourly and IR_hourly) were visualized for each individual and monitoring week (Tuckey plots). Hourly values that exceeded 1.5*IQR of the distribution for a specific hour of the day within each monitoring week was excluded. This was done to eliminate extreme values that may not be representative of long- to mid-term community noise exposure at a given location and time.

###### "simple outlier exclusion 

```{r message=FALSE, warning=FALSE, include=FALSE}
library(pdftools)
show_pdf_firstpage <- function(pdf, pages = 1:5) {

  tmp <- tempfile(fileext = ".pdf")

  pdftools::pdf_subset(
    pdf,
    pages = pages,
    output = tmp
  )

  # IMPORTANT: use embed, not iframe
  htmltools::browsable(
    htmltools::tags$embed(
      src = tmp,
      type = "application/pdf",
      width = "100%",
      height = "650px"
    )
  )
}
```


```{r echo=FALSE, out.width="100%"}
show_pdf_firstpage(
    "../data processing/noise_reports/noiseplots_outlier_week_2.pdf",
  pages = 1:3
)

```

###### 1h Tuckey IQR exclusion


```{r echo=FALSE, out.width="100%"}
show_pdf_firstpage(
    "../data processing/noise_reports/noiseplots_outlier_corrected_LAeqweek_2.pdf",
  pages = 1:3
)

```


```{r 2a daily plots, echo=FALSE, fig.height=3, fig.width=8, fig.align='center'}
ggplot(data = noise_ind_daily, aes(x = dow)) +
  geom_line(aes(y = L_Aeq_tot, color = "L_Aeq 24h")) +
  geom_line(aes(y = L_Aeq_day, color = "L_Aeq day")) +
  geom_line(aes(y = L_Aeq_night, color = "L_Aeq night"))  +
  geom_line(aes(y = L_den, color = "L_den")) +
  scale_color_manual(values = c("L_Aeq 24h" = "green4", 
                                "L_Aeq day" = "gold", 
                                "L_Aeq night" = "grey2",
                                "L_den" = "red")) +
  theme_minimal() +
  labs(x = "Day of the week", y = "L_Aeq [dB]", color = "Legend") +
  lims(y = c(40,70))


ggplot(data = noise_ind_daily, aes(x = dow)) +
  geom_line(aes(y = IR_24h, color = "IR 24h")) +
  geom_line(aes(y = IR_day, color = "IR day")) +
  geom_line(aes(y = IR_night, color = "IR night"))  +
  scale_color_manual(values = c("IR 24h" = "green4", 
                                "IR day" = "gold", 
                                "IR night" = "grey2")) +
  theme_minimal() +
  labs(x = "Day of the week", 
       y = "Intermittency ratio [%]", 
       color = "Legend") +
  lims(y = c(0, 250))
```

<br>



#### Summary

This cleaning protocol has been completed and replicated throughout all 4 monitoring weeks.


<br>
